---
title: "Campaign Contributions EDA Project"
author: "Grigory Makarenko"
output:
  html_document: default
  html_notebook: default
---

========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      message=FALSE,
                      warning=FALSE,
                      fig.width=10,
                      fig.height=7)
```

```{r packages}
# Load packages

if (!require(plyr)) install.packages('plyr')
library(plyr)

if (!require(dplyr)) install.packages('dplyr')
library(dplyr)

if (!require(tidyr)) install.packages('tidyr')
library(tidyr)

if (!require(ggplot2)) install.packages('ggplot2')
library(ggplot2)
theme_set(theme_bw())

if (!require(zoo)) install.packages('zoo')
library(zoo)
```

```{r}
# Load the data set
df <- read.csv("virginia_contributions_mod.csv", header=TRUE, sep=",")
```

Presidential Map Contributor Dataset from the Federal Election Commission is a dataset of all individual campaign contributions, refunds to individuals, and transfers from authorized committees. This report will explore a subset of this dataset for contributions from Virginia during the 2016 election cycle. 

# Univariate Plots

```{r}
dim(df)
```

```{r}
str(df)
```

This dataset had 20 features and 212849 observations. Most of the features are factor type. 

There are 25 different candidates and committees. Contributions to a candidate go through their committee, so they are essentially the same variable. Each candidate/committee is associated with one of 5 parties (in this dataset).

Some of the features have to be adjusted before analysis. The contb_receipt_dt feature needs to be a date type. The contbr_zip is a num type but it would make more sense as a factor type. It might be helpful to convert any zip-plus-4 codes to normal zip codes.

Some features can be removed because they would not contribute to the analysis. cand_id, cmte_id, file_num, and contbr_st do not add any extra information from an analysis perspective.
It is interesting that while tran_id is a unique id it has slightly less levels than the number of observations.

```{r}
# Remove unneeded features
df <- subset(df, select = -c(cand_id, cmte_id, file_num, contbr_st))

# Convert features to correct data type
df$contbr_zip <- factor(df$contbr_zip)
df$contbr_zip_p4 <- df$contbr_zip

df$contbr_zip <- sapply(df$contbr_zip,
                        FUN = as.character)

df$contbr_zip <- sapply(df$contbr_zip,
                        FUN = function(x) substr(x, start=1, stop=5))

df$contbr_zip <- factor(df$contbr_zip)

df$contb_receipt_dt <- as.Date(df$contb_receipt_dt, "%d-%b-%y")
```

```{r}
summary(df)
```

We can see that the contribution receipt dates range 3 years from 2013-12-31 to 2016-12-31. It's clear that the committees matched to the candidate. Democratic candidates received almost 3 times the number of contributions of the other parties combined. Also, Hillary Clinton got more than 50% of all the contributions in the dataset. Did she also get more than 50% of the aggregate contribution amount? 
The top 5 contribution counts are from the Northern Virginia area. This is not surprising because of the population density and proximity to Washington, DC. Most contributions do not name an employer, and a 1/4 of the contributions are from retired people. The majority of receipts had no description.

## Timeline {.tabset}

### Daily
```{r}
date.count <- df %>%
  count(contb_receipt_dt)
```

```{r}
ggplot(data = date.count, aes(x = contb_receipt_dt, y = n)) +
  geom_point() +
  geom_smooth()
```

From 2014 to 2015 there were few contributions to the national campaign with many days completely free of contributions. 2015 to 2016 saw a steady incline of the frequency of contributions, but the daily numbers stayed below 500 contributions. 2016 was scattered mix with many spikes in contributions. These spikes can likely be explained by major events or scandals that occurred during this election period.

### Monthly Timeline
```{r}
month.count <- df %>%
  mutate(contb_receipt_dt = as.Date(as.yearmon(contb_receipt_dt), frac=1)) %>%
  count(contb_receipt_dt)
```
```{r}
horiz_lab <- theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
```
```{r}
ggplot(data = month.count, aes(x = contb_receipt_dt, y = n)) +
  geom_bar(stat="identity") +
  horiz_lab
```

With this division, we can plainly the spikes in contributions by month. We can see that the first uptick was during the primary election from February 2016 to April 2016. Then we have decline followed by big increase in July 2016 leading up to the presidential election in November 2016. October 2016 was the last push to campaign before election day, and it shows in the major increase. It's interesting that November still had so many contributions even when the election date is so early in the month.

##

## Candidates {.tabset}

### Candidate Summary
```{r}
df$cand_nm <- sapply(df$cand_nm, FUN = as.character)

cand.count <- df %>%
  select(cand_nm) %>%
  count(cand_nm) %>%
  arrange(desc(n))
```
```{r}
cand.count
```

### All contributions

```{r}
ggplot(data = cand.count, aes(x = reorder(cand_nm, -n), y = n)) +
  geom_bar(stat="identity") +
  horiz_lab
```

### Less than 15000 contributions

```{r}
ggplot(data = cand.count[cand.count$n < 15000,],
       aes(x = reorder(cand_nm, -n), y = n)) +
  geom_bar(stat="identity") +
  horiz_lab
```

### Less than 5000 contributions

```{r}
ggplot(data = cand.count[cand.count$n < 5000,],
       aes(x = reorder(cand_nm, -n), y = n)) +
  geom_bar(stat="identity") +
  horiz_lab
```

### Less than 500 contributions

```{r}
ggplot(data = cand.count[cand.count$n < 500,],
       aes(x = reorder(cand_nm, -n), y = n)) +
  geom_bar(stat="identity") +
  horiz_lab
```

### Party Aggregate

```{r}
ggplot(data = df, aes(x = cand_party)) +
  geom_bar() +
  scale_y_sqrt()
```
The y axis was scaled by square root to be able to see the low contribution parties.

##
We can see that most of the candidates got very few contributions compared to Clinton, Sanders, and Trump. Even if the top 3 "outliers" are removed, we can see that most of the 25 candidates received less than 5000 contributions. More than half of the candidates received even less than 250 contributions. It will be interesting to see how the number of contributions relates to the amount contributed. Republicans could be getting less contribution but those contributions have a high receipt amount. So, are the aggregate receipt amounts more even between the parties?

## Geographic Data Graphs {.tabset}

### Cities (Top 20 Count)

```{r}
# help from https://www.r-bloggers.com/data-frame-columns-as-arguments-to-dplyr-functions/

df$contbr_city = sapply(df$contbr_city, FUN = as.character)

reduce_count <- function(x, y) {
  x %>%
    select_(y) %>%
    count_(y) -> new_out
  return(new_out)
}

city.count <- reduce_count(df, "contbr_city")
```
```{r}
ggplot(data = top_n(city.count, 20, n),
       aes(x = reorder(contbr_city, -n), y = n)) +
   geom_bar(stat="identity") +
   horiz_lab
```

Most of the cities in the top 20 are near Washington, DC. The other ones (Richmond, Virginia Beach) are major population centers. These results aren't surprising. It is interesting that Henrico is actually the county surrounding Richmond. It is clear that there are some anomalies in the dataset.

We will see that better by looking a the low count cities.

### Cities (Low Counts)

```{r}
subset(city.count, city.count$n == 1)
```

There are 265 rows that had a count of 1. Many of these are misspelled names or names of cities outside of Virginia (New York, Las Vegas). While the misspellings could be cleaned up, it doesn't seem worth the effort for such a low amount relative to the whole sample size.

### Zipcodes

```{r}
zip.count <- reduce_count(df, "contbr_zip")
```
```{r}
ggplot(data = top_n(zip.count, 40, n),
       aes(x = reorder(contbr_zip, n), y = n)) +
  geom_bar(stat="identity") +
  coord_flip(expand=TRUE)
```

The amount of contributions per zip code looks close to the cities with a few more spikes. Some cities contain several zip codes so much of the spread is just subdivision of the city amounts.

### Zipcodes Distribution

```{r}
ggplot(data = zip.count, aes(x = n)) +
  geom_histogram(bins = 100) +
  scale_x_sqrt() +
  labs(y = "Number of Zipcodes", x = "Number of Contributions")
```

This plot displays how many zipcodes (locations) had certain numbers of contributions, e.g. only two zipcode had more than 5000 contributions.

I scaled the plot by a square root. The histogram shows that most of the zipcodes that were reported in the dateset had less than 500 contributions. Meaning that most of the contributions came from a few zipcodes (or geographic areas).

##

## Contributor Employment {.tabset}

### Contributor Occupation
```{r}
# many of these changes felt necessary only after getting futher into the analysis
df$contbr_occupation <- sapply(df$contbr_occupation,
                               FUN = as.character)

df$contbr_occupation <- gsub("^INFORMATION REQUESTED PER BEST EFFORTS$",
                             "INFORMATION REQUESTED",
                             df$contbr_occupation)

df$contbr_occupation <- gsub("^LAWYER$", "ATTORNEY",
                             df$contbr_occupation)

occupation.count <- reduce_count(df, "contbr_occupation")
```
```{r}
ggplot(data = top_n(occupation.count, 20, n),
       aes(x = reorder(contbr_occupation, n), y = n)) +
  geom_bar(stat="identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size=8))
```

The majority of the top 20 contributors were retired. It's interesting that the next highest group said they were unemployed. It will be interesting to see how much each occupation group actually contributed. "Information requested" and "Information requested per best efforts" are best effort designations showing that the committee tried to get additional information from a contributor. 

### Contributor Employer
```{r}
# many of these changes felt necessary only after getting further into the analysis
df$contbr_employer <- sapply(df$contbr_employer,
                             FUN = as.character)

df$contbr_employer <- gsub("^INFORMATION REQUESTED PER BEST EFFORTS$",
                           "INFORMATION REQUESTED",
                           df$contbr_employer)

df$contbr_employer <- gsub("^SELF$", "SELF-EMPLOYED",
                           df$contbr_employer)

df$contbr_employer <- gsub("^SELF EMPLOYED$", "SELF-EMPLOYED",
                           df$contbr_employer)

df$contbr_employer <- gsub("^DOD$", "U.S. DEPARTMENT OF DEFENSE",
                           df$contbr_employer)

df$contbr_employer <- gsub("^US NAVY$", "U.S. NAVY",
                           df$contbr_employer)

employer.count <- reduce_count(df, "contbr_employer")
```
```{r}
ggplot(data = top_n(employer.count, 20, n),
       aes(x = reorder(contbr_employer, n), y = n)) +
  geom_bar(stat="identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size=8))
```

Many of the contributors chose to not give their employer name. Again, many people are retired or have no employer. Interestingly many people report self-employment.

##

## Contribution Receipt Amount {.tabset}

### Raw Histogram

```{r}
receipt_amt_graph <- ggplot(data = df, aes(x = contb_receipt_amt)) +
  geom_histogram(binwidth=100)
```
```{r}
receipt_amt_graph
```

As the summary showed, the median contribution receipt amount is within \$100 or \$37 to be exact.

### Zoomed Histogram

```{r}
receipt_amt_graph +
  scale_y_sqrt() +
  coord_cartesian(ylim = c(0, 50000))
```

I scaled the counts by a square root to be able to see the lower count receipt amounts. Looking closer we can see that small contributions are the most popular, which isn't particularly surprising. Although, there are spikes at roughly \$300 and \$2500. There are also a few negative amounts or returned contributions.

##

# Univariate Analysis

The dataset started with 20 features, several of them redundant. 212849 contributions were made within a span of 3 years from December 31, 2013 to December 31, 2017. Some of these logged contributions were actually returns. About 3/4 of the contributions went to Democratic candidates/committees. Most of the contributions come from the Northern Virginia area and major population centers. The vast majority of contributions are within \$100.

The committee name and political party features was added to this dataset through a separate python script. Although, the committee features do not give any more information than the candidate name feature.

The zip codes feature was reduced to only 5 digits. Zip code + 4 is very specific, and the change makes it more broad. The receipt date feature had to be changed from a factor type to an actual Date type.

Several features could be cleaned up, but that is unnecessary. For example, we saw that there were several cities that were misspelled or were not Virginia cities. That amount is negligible relative to the dataset, so the effort of cleaning the data would be much higher than the value of the extra datapoint.

```{r}
city.count.count <- city.count %>%
  mutate(nf = factor(n)) %>%
  select(nf) %>%
  count(nf)
```
```{r}
head(city.count.count, 10)
```

For the city feature, I will assume that the city names with counts of up to 10 (nf means factor of the count) would have some data that would require cleaning. The count of each of those is above.

```{r}
city.count.count %>%
  # Learned about the way to avoid information loss on conversion from factor to
  # numeric at the link below.
  # http://stackoverflow.com/questions/3418128/how-to-convert-a-factor-to-an-integer-numeric-without-a-loss-of-information
  mutate(n_product = as.numeric(levels(nf))[nf] * n) %>%
  summarise(sum(head(n_product, 10)))
```

If I take the sum of the products of each of the 10 rows, I can get a rough idea about the amount noisy datapoints we might have (above). So that's 0.74982734238% of the data points. I think other variables (ie. employer, occupation) with free text input would be similar.

I want to see which variables might affect 1) the candidate or party that is contributed to and 2) the receipt amount of the contribution. The geographic and occupation are the only demographic variables for contributors, so they might not provide much more than basic explanatory power.


# Bivariate Plots

## Receipt Amount over Time {.tabset}

```{r}
date.receipts.summary <- df %>%
  mutate(contb_receipt_dt = as.Date(as.yearmon(contb_receipt_dt), frac=1)) %>%
  group_by(contb_receipt_dt)
```

### Receipts Boxplots
```{r}
ggplot(data = subset(date.receipts.summary,
                     contb_receipt_amt > 0), 
       aes(x = contb_receipt_dt,
           y = contb_receipt_amt,
           group = contb_receipt_dt)) +
  geom_boxplot(outlier.shape = 1) +
  scale_y_log10()
```

The contibution receipt amount was scaled by a log. Negative receipts were removed.

The mean and median contribution receipt amounts throughout the election are quite interesting. They are almost the opposite of the frequency of contributions in the same time period. Earlier in the election there were fewer contributions per month, so the means are much more heavily affected by any large contribution. This plot is indicating that there are outliers in the data. I do not expect there contributions to be actual outliers, rather they are just very rare for each month's contribution receipt amounts.

### Aggregate Receipts
```{r}
ggplot(data = date.receipts.summary %>% 
         summarise(agg_receipt = sum(contb_receipt_amt)),
       aes(x = contb_receipt_dt)) +
  geom_bar(aes(y = agg_receipt), stat="identity") +
  horiz_lab
```

The aggregate receipts per month follow much closer to the contribution counts per month. We can see again that many people contributed around the party primary elections and the general election.

##

## Amount by Candidate {.tabset}

### Receipts Boxplots
```{r}
ggplot(data = subset(df,
                     contb_receipt_amt > 0), 
       aes(x = reorder(cand_nm, -contb_receipt_amt, mean),
           y = contb_receipt_amt,
           group = cand_nm)) +
  geom_boxplot(outlier.shape = 1) +
  scale_y_log10() +
  horiz_lab
```

The contibution receipt amount was scaled by a log. Negative receipts were removed. Ordered by mean.

The two democratic candidates that got the most contributions have some of the lowest median and mean receipts. Bernie Sanders's mean is the lowest out of the group. Ted Cruz and Donald Trump had the largest value receipts, and Hillary Clinton had the lowest value positive receipt.

### Aggregate Receipts
```{r}
ggplot(data = df %>%
         group_by(cand_nm) %>%
         summarise(agg_receipt = sum(contb_receipt_amt)),
       aes(x = reorder(cand_nm, -agg_receipt), y = agg_receipt)) +
  geom_bar(stat="identity") +
  scale_y_sqrt() +
  horiz_lab
```
The y axis was scaled by a square root.

Hillary Clinton received the highest amount of money from Virginia contributors by a large margin. It's surprising that Jeb Bush had a similar aggregate receipt amount as Ted Cruz, but Bush had about 1/10 of the number of contributions as Cruz. Similarly, Bernie Sanders had close to twice the number of contributions as Donald Trump but had half of the aggregate receipt amount.

### Receipt Distribution
```{r}
top_4_cand <- top_n(cand.count, 4, n)["cand_nm"]
```
```{r}
ggplot(data = df %>% inner_join(top_4_cand), aes(x = contb_receipt_amt)) +
  geom_histogram(binwidth = 50) +
  scale_y_sqrt() +
  coord_cartesian(xlim = c(-5000, 5000), ylim = c(0, 20000)) +
  facet_wrap(~ cand_nm)

```

Looking at the receipt distribution of the top 4 candidates (by count), we can see that for each candidate the distribution is heavily centered at contributions that are less than \$500. Hillary Clinton's graph has been reduced, but it reaches up to 50000 contributions in the center. It's hard to tell whether Ted Cruz's or Bernie Sanders's distribution is the most narrow of the 4 cardidates.

### Summary Assessment

It's clear that Hillary Clinton was the most popular candidate in Virginia. She had the most contributions and did not get the highest aggregate receipts just by high value donations. 
It isn't necessary to look at the impact of political party on the number of contributions or the overall receipts, since all the republican candidates combined wouldn't overtake Hillary Clinton's aggregate receipts.

##

## Receipt Amount by Location {.tabset}

```{r}
top_30_city <- top_n(city.count, 30, n)["contbr_city"]
```

### Receipts Boxplots
```{r}
ggplot(data = subset(df, contb_receipt_amt > 0) %>% inner_join(top_30_city), 
       aes(x = reorder(contbr_city, -contb_receipt_amt, mean),
           y = contb_receipt_amt,
           group = contbr_city)) +
  geom_boxplot(outlier.shape = 1) +
  scale_y_log10() +
  horiz_lab
```

The contibution receipt amount was scaled by a log. Negative receipts were removed. Ordered by mean.

McLean has a significantly higher mean receipt than the rest of the top 30 contributing cities. However, McClean had a significantly lower contribution count than Arlington and Alexandria. It is not surprising that the top 5 cities in this group are in the Northern Virginia area which is, on average, the richest area in Virginia and a Washington, DC suburb. The biggest contribution came from a person in Alexandria, and the rest of the cities' contributions were quite a bit lower maximum contribution receipt. 

### Aggregate Receipt
```{r}
ggplot(data = df %>%
         group_by(contbr_city) %>%
         summarise(agg_receipt = sum(contb_receipt_amt)) %>%
         inner_join(top_30_city),
       aes(x = reorder(contbr_city, -agg_receipt), y = agg_receipt)) +
  geom_bar(stat="identity") +
  scale_y_sqrt() +
  horiz_lab
```

The y axis was scaled by a square root.

As would be expected, Arlington and Alexandria have the highest aggregate contribution receipts. McLean surpassed Richmond, but it had close to half as many contributions as Richmond. McLean did have a much higher mean receipt than every other city.

### Summary Analysis

Generally, the population centers of Virginia are represented on the list of top 30 contribution counts. We can see that Northern Virginia cities dominate the list of the highest contribution amounts and receipt amounts. It isn't completely surprising since Northern Virginia is close to DC, and the people in this area will be the most affected by a national election. McLean is an interesting city because it had such a high mean receipt amount from around 6000 contributions. How are these contributions distributed between the candidates or parties? We haven't looked much at the actual contributors by name. Could some of these statistics be influenced by just a few individuals?

##

## Receipts by Occupation {.tabset}

```{r}
top_30_occupation <- top_n(occupation.count, 30, n)["contbr_occupation"]
```

### Receipts Boxplots
```{r}
ggplot(data = subset(df, contb_receipt_amt > 0) %>% inner_join(top_30_occupation),
       aes(x = reorder(contbr_occupation, -contb_receipt_amt, mean),
           y = contb_receipt_amt,
           group = contbr_occupation)) +
  geom_boxplot(outlier.shape = 1) +
  scale_y_log10() +
  horiz_lab
```

The contibution receipt amount was scaled by a log. Negative receipts were removed. Ordered by mean.

Executives have a high mean and median contribution receipt amount which is not unusual. Homemakers had a surprisingly high mean receipt amount, maybe related to executives in some way. It is also interesting that RNs had a lower mean receipt than even unemployed people. Are RNs generally less likely to contribute large amounts or is this particular to this election or state?

The maximum receipts look more even between groups than was the mean. Again, RN's are at the bottom of the list with a significantly lower maximum contribution than the rest of the groups.

### Aggregate Receipt
```{r}
ggplot(data = df %>%
         group_by(contbr_occupation) %>%
         summarise(agg_receipt = sum(contb_receipt_amt)) %>%
         inner_join(top_30_occupation),
       aes(x = reorder(contbr_occupation, -agg_receipt), y = agg_receipt)) +
  geom_bar(stat="identity") +
  scale_y_sqrt() +
  horiz_lab
```

The y axis was scaled by a square root.

Overall, retired people contributed \$5000000 of the \$28000000 contributed this campaign. The amount quickly declines after retirees. Homemakers and unemployed people surpassed executives in aggregate contributions. Is the frequency of contributions related to the average amount of individual leisure time? That would help explain the prominence of some of the groups in this dataset.

### Summary Analysis

The contribution receipts by occupation had some interesting results. We saw from looking at the counts that the most frequent contributions came from retirees, unemployed, info requested, attorneys, and consultants, in that order. While there were around 13000 contributions from the unemployed group, their average contribution was quite small compared to the rest of the top 5 groups. Surprisingly homemakers had a higher aggregate receipt amount with only 3000 contributions.

##

## Receipts by Employer {.tabset}

```{r}
top_30_employer <- top_n(employer.count, 30, n)["contbr_employer"]
```

### Receipts Boxplots
```{r}
ggplot(data = subset(df, contb_receipt_amt > 0) %>% inner_join(top_30_employer),
       aes(x = reorder(contbr_employer, -contb_receipt_amt, mean),
           y = contb_receipt_amt,
           group = contbr_employer)) +
  geom_boxplot(outlier.shape = 1) +
  scale_y_log10() +
  horiz_lab
```

The contibution receipt amount was scaled by a log. Negative receipts were removed. Ordered by mean.

The mean receipt of homemakers in the occupation column is slightly different than the one for employer. It could be that some of the people that listed themselves as a homemaker could have left the employer section blank or written self-employed. Out of the `r occupation.count$n[occupation.count$contbr_occupation == "HOMEMAKER"]` that listed themselves as homemaker, `r length(df$contbr_occupation[df$contbr_employer == "HOMEMAKER"])` also listed homemaker on employer. It's interesting how many actual companies are represented in the dataset.

The maximum contribution is even between most of the employers. The self-employed and the blank groups could contain so many different types of individuals that it's hard to get anything out of their maximum contributions.

The "TEST" group is hard to determine. It might be that some of contributions in this dataset were just test to test the system.

### Aggregate Receipt
```{r}
ggplot(data = df %>%
         group_by(contbr_employer) %>%
         summarise(agg_receipt = sum(contb_receipt_amt)) %>%
         inner_join(top_30_employer),
       aes(x = reorder(contbr_employer, -agg_receipt), y = agg_receipt)) +
  geom_bar(stat="identity") +
  scale_y_sqrt() +
  horiz_lab
```

The y axis was scaled by a square root.

It isn't surprising that the biggest groups of contributors would have the highest aggregate receipt amount. It seems that there are too many different employers to be able to gain too much information about the contributions. It is clear that large organizations or groups (retiree) will be more respresented in this list. 

### Summary Analysis

The most obvious thing about the employer feature is that it is too broad. There are also many names that could be fixed or consolidated. While I was able to convert a few of the major employers (DOD, US Navy), more issues become apparent. Like I discussed earlier about cities, it might not be worth the effort to try to clean up this variable.

##

## Candidate Contributions by Date {.tabset}

### All Candidates
```{r}
cand.date.count <- df %>%
  select(cand_nm, contb_receipt_dt) %>%
  mutate(contb_receipt_dt = as.Date(as.yearmon(contb_receipt_dt), frac = 1)) %>%
  group_by(contb_receipt_dt) %>%
  count(cand_nm)
```
```{r}
ggplot(data = cand.date.count,
       aes(x = contb_receipt_dt, y = n, fill = cand_nm)) +
  geom_bar(stat = "identity")
```

### Top 5 Candidates
```{r}
top_5_cand <- top_n(cand.count, 5, n)["cand_nm"]
```
```{r}
ggplot(data = cand.date.count %>% inner_join(top_5_cand),
       aes(x = contb_receipt_dt, y = n)) +
  geom_bar(stat = "identity") +
  facet_wrap(~cand_nm) +
  horiz_lab
```

##

These plots paint an interesting picture of the election, at least in Virginia. The Virginia Party Primary elections were is March 2016. Looking at the contributions leading up to March, I would expect Bernie Sanders to win the Democratic delegates and a close race between the 2 Republican delegates. What we actually saw was a substantial win from Hillary Clinton and a surprising win from Donald Trump. Donald Trump had barely any contributions even up to July 2016, when the Primary elections ended. A similar thing seems to have occured during the General election. The 2 Presidential candidates had a close race in Virginia, but the contribution numbers would make me expect a landslide win by Hillary Clinton.

# Bivariate Analysis

In this part of the analysis I kept most of my attention on the monetary amounts that were contributed during this election. We can see that most of the earlier contributions might be coming from people that were more informed about the candidates that would be running in this election. I think that because the early donations had a fairly large mean. As more people started to get interested in the campaign, their smaller donations quickly brought down the averages. 

When looking at contributions by city, I didn't find it surprising that the most populated areas in Virginia would have the highest frequency of contributions. It was interesting to see some of the smaller towns have high numbers. McClean has a population of about 48 thousand people (in 2010); 5.8 thousand contributions came from this area and equaled to $2.36 million. That impressive when we consider that Alexandria's population is 3 times as large with 4 times the contribution count but only 2 times the aggregate receipt amount.

It is clear that the contribution rates and amounts are affected by some events that are outside of the dataset. However, there seems to be some separation between contribution trends and actual voting trends. As I stated earlier, I would not have expected Donald Trump to be in the race at all if I looked at his contribution numbers before the Primary Election. Nor would I expect a close race in Virginia between Clinton and Trump.

# Multivariate Plots

## Location/Candidates Receipts {.tabset}

```{r}
city.receipts.cand <- df %>%
  select(contbr_city, cand_nm, cand_party, contb_receipt_amt) %>%
  group_by(cand_nm) %>%
  inner_join(top_30_city) %>%
  filter(contb_receipt_amt >= 0)
```

### Top 30 Cities/All Candidates
```{r}
ggplot(data = city.receipts.cand, aes(x = contbr_city, y = contb_receipt_amt, fill = cand_nm)) +
  geom_bar(stat="identity") +
  horiz_lab
```

As would be expected, Hillary Clinton's contribution receipt amount from each city overshadows the rest of the candidates.

### Top 30 Cities/All Candidates - HRC
```{r}
ggplot(data = city.receipts.cand %>%
         filter(!cand_nm %in% c("Clinton, Hillary Rodham")), 
       aes(x = contbr_city, y = contb_receipt_amt, fill = cand_nm)) +
  geom_bar(stat="identity") +
  horiz_lab
```

Looking at the chart with HRC helps to see that Jeb Bush actually received a relatively high contribution receipt amount from the highest contributing cities. That's quite unexpected because he fell out of the campaign quite early.

### Top 30 Cities/Political Party
```{r}
ggplot(data = city.receipts.cand %>%
         filter(cand_party %in% c("REPUBLICAN", "DEMOCRATIC")),
       aes(x = contbr_city, y = contb_receipt_amt, fill = cand_party)) +
  geom_bar(stat="identity") +
  horiz_lab
```

Few cities of the 30 highest contributing cities actually gave more to Republicans than Democrats. That is an obvious fact based on Hillary Clinton's receipts. The question would be whether these campaign contributors are Democratic Party supporters or Hillary Clinton supporters.

### Top 30 Cities/Political Party - HRC
```{r}
ggplot(data = city.receipts.cand %>% 
         filter(cand_party %in% c("REPUBLICAN", "DEMOCRATIC")) %>% 
         filter(!cand_nm %in% c("Clinton, Hillary Rodham")),
         aes(x = contbr_city, y = contb_receipt_amt, fill = cand_party)) +
  geom_bar(stat="identity") +
  horiz_lab
```

Removing HRC from the included Democrats really shows how much of the overall contributions went to her campaign rather than Democratic candidates in general. There can be many reasons for this type of contibution strategy, but I would suggest that one of the major reasons is the small pool of candidates that ran in the Democratic Primary Election.

##

## Date/Candidate Receipts {.tabset}

### Date/All Candidates
```{r}
date.receipts.cand <- df %>%
  select(contb_receipt_dt, cand_nm, cand_party, contb_receipt_amt) %>%
  mutate(contb_receipt_dt = as.Date(as.yearmon(contb_receipt_dt), frac = 1)) %>%
  group_by(cand_nm) %>%
  filter(contb_receipt_amt >= 0)
```

```{r}
ggplot(data = date.receipts.cand,
       aes(x = contb_receipt_dt, y = contb_receipt_amt, fill = cand_nm)) +
  geom_bar(stat="identity") +
  horiz_lab
```

When you compare this chart with the one showing contribution frequencies, you can see a difference between the number of contributions some of the candidates recieve and the aggregate dollar amount of those contributions. In early 2016 Bernie Sanders had a clear lead in the amount of contribution he recieved, but that lead is not reflected in the actual dollar value of those contributions.

### Date/Top 5 Candidates
```{r}
date.receipts.cand.t5 <- date.receipts.cand %>%
  inner_join(top_5_cand)
```

```{r}
ggplot(data = date.receipts.cand.t5,
       aes(x = contb_receipt_dt, y = contb_receipt_amt)) +
  geom_bar(stat="identity") +
  facet_wrap(~ cand_nm) +
  horiz_lab
```

It's interesting to clearly see the switch in funding after the Primary Election was finished. After Donald Trump won the Primary, his contribution receipts jumped substantially. The effect was even greater for Hillary Clinton. Also worth noting is the increase in funds does not look like just a shift from the other candidates. It seems that contributors give higher amounts after each party's candidate is selected.

### Date/Polical Parties
```{r}
ggplot(data = date.receipts.cand,
       aes(x = contb_receipt_dt, y = contb_receipt_amt, fill = cand_party)) +
  geom_bar(stat="identity") +
  horiz_lab
```

The timeline for the contributions to each party have roughly two peaks. The first is some time around early 2016, possibly due to candidates dropping out of the race and the uncertainty that creates for potential contributors. Then the next peak is around the General Election day. 

### Date/Party/Candidates

#### Democrats
```{r}
ggplot(data = filter(date.receipts.cand, cand_party == "DEMOCRATIC"),
       aes(x = contb_receipt_dt, y = contb_receipt_amt, fill = cand_nm)) +
  geom_bar(stat="identity") +
  horiz_lab
```

#### Republicans
```{r}
ggplot(data = filter(date.receipts.cand, cand_party == "REPUBLICAN"),
       aes(x = contb_receipt_dt, y = contb_receipt_amt, fill = cand_nm)) +
  geom_bar(stat="identity") +
  horiz_lab
```

The parties adopted quite different strategies for their election process. The Democratic had 4 candidates to the 17 Republican candidates. The Democratic chart looks like it Hillary was always going to be the Democratic nominee. The Republican chart has a lot of variability in the candidate with the most contributions until after the Primary election was completed. However, this dataset only represents the population that contributed the candidates, and there might not be any correlation between this population and the general voting population.

##

# Multivariate Analysis

Looking at the data through a temporal lense seems to be most interesting. The trend of contribution during this election have 2 peaks. This is evident for both parties in varying degrees. The first peak is associated with the buildup to the parties' Primary Elections. The second peak is associated with the General Election. The Democratic party seems to have a more stable Primary Election phase, with only a few possible candidates to choose. The Republican party has a more open nominee selection process.

Separating the candidates into their own graphs gives a much clearer picture of the difference between the funding provided to each candidate. As a broader theme, the amount of funding a candidate receives does not seems to closely match with the way people actually vote. Several of the results of the Virginia elections do not match what we see in these charts.

# Final Plots and Summary

## Plot One

```{r}
ggplot(data = date.count, aes(x = contb_receipt_dt, y = n)) +
  geom_point() +
  geom_smooth(span = 0.4) +
  labs(title = "All Contributions per Day", x = "Date", y = "Count")
```

## Description One

This plot shows the contribution frequency by date for the election period. The local regression line provides a nice trend line for the counts. Most of the data instances are dated after early 2015. The confidence interval around the trend line quickly shrunk after the rate of contribution increased in early 2015. 

## Plot Two

```{r}
ggplot(data = subset(date.receipts.summary,
                     contb_receipt_amt > 0), 
       aes(x = contb_receipt_dt,
           y = contb_receipt_amt,
           group = contb_receipt_dt)) +
  geom_boxplot(outlier.shape = 1) +
  scale_y_log10() +
  labs(title = "Distribution of Receipts per Month",
       x = "Date",
       y = "log(Receipt Amount [USD])")
```

## Description Two

This boxplot of receipts per month shows a difference between the contributions that the candidates got throughout the election process. Both the mean and the median is relatively high earlier in the election. As the candidates get further into the election, the mean and median condibution receipt amount decreases until the late election. A lot more "outliers"  are observed during the most active months of the election. While they might be interesting to investigate further, I do not think most of these values are outliers because of errors.

## Plot Three

```{r}
ggplot(data = filter(date.receipts.cand, cand_party == "DEMOCRATIC"),
       aes(x = contb_receipt_dt, y = contb_receipt_amt, fill = cand_nm)) +
  geom_bar(stat="identity") +
  horiz_lab +
  labs(title = "Receipts per Month by Democratic Candidate",
       x = "Date",
       y = "Receipt Amount [USD]")
```

```{r}
ggplot(data = filter(date.receipts.cand, cand_party == "REPUBLICAN"),
       aes(x = contb_receipt_dt, y = contb_receipt_amt, fill = cand_nm)) +
  geom_bar(stat="identity") +
  horiz_lab +
  labs(title = "Receipts per Month by Republican Candidate",
       x = "Date",
       y = "Receipt Amount [USD]")
```

## Description Three

While this is two plots, they are interrelated. The difference between and within the candidates is the most important aspect of this dataset. On the Democrats side Hillary Clinton had more funding throughout the whole election cycle in Virginia. Also, the competition was mostly between Hillary Clinton and Bernie Sanders. The Republican side looks very different. There were many more candidates represented in the election, and it was not evident who was the favorite until after the Primary Election. These plots together provide a clear distinction between the Republican and Democratic Primary election more than the General election.

# Reflection

In my exploration of the 2016 US presidential election in Virginia dataset, I always thought something was missing.

For this exploration, I decided that looking at the features that seemed most likely to be interesting. There are features that have a very specific accounting or compliance purpose. There are features that serve an indentification purpose, such as contributor names. I chose to just skip features that had a compliance purpose or could not be even partially generalizable. I had to create a feature for the candidates' party and committee names. I used the Federal Election Commission website to find the candidates' parties and committee names by ID. A short python script created the features and export a new csv.

Most of the features I used can have misspellings and missing values or are completely misleading. If I was planning to do more work on this dataset, I would have put much more effort into fixing the misspelling. Like I mentioned in the exploration, the effort to benefit ratio will constitute how much time should be allocated to cleaning the dataset. Figuring out how to deal with missing values or impossible values (e.g. Hong Kong for city) is more difficult and would depend on the use case. In most cases, I would think that missing values should just be removed from analysis since most of the time we are dealing with contributors' free-text answers. However, creating new features for contributors' demographics might have some insightful results. This could be things like contibutors' occupation industry rather than specific occupation or residential counties rather cities or zipcode.

I think my greatest challenge was understanding what this dataset actually represents. Throughout the exploration I wanted to consider how these features could explain the results of the 2016 election in Virginia. However, making that leap would require certain assumptions or insights about how people contribute to elections versus how they actually vote. I think that's a very interesting topic to explore, but I doubt this dataset alone would provide a sufficient or generalizable explanation. This dataset only explains how people contributed in Virginia in during the 2016 election.
