---
title: "Campaign Contributions EDA Project"
author: "Grigory Makarenko"
output:
  html_document: default
  html_notebook: default
---

========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r packages}
# Load packages

if (!require(plyr)) install.packages('plyr')
library(plyr)

if (!require(dplyr)) install.packages('dplyr')
library(dplyr)

if (!require(tidyr)) install.packages('tidyr')
library(tidyr)

if (!require(ggplot2)) install.packages('ggplot2')
library(ggplot2)
theme_set(theme_bw())

if (!require(GGally)) install.packages('GGally')
library(GGally)

if (!require(corrplot)) install.packages('corrplot')
library(corrplot)
```

```{r load_data}
# Load the data set
df <- read.csv("virginia_contributions.csv", header = TRUE, sep = ",")
```

Presidential Map Contributor Dataset from the Federal Election Commission is a dataset of all individual campaign contributions, refunds to individuals, and transfers from authorized committees. This report will explore a subset of this dataset for contributions from Virginia during the 2016 election cycle. 

##Univariate Analysis

```{r data_dim}
dim(df)
```

```{r data_str}
str(df)
```

This dataset had 18 features and 212849 observations. Most of the features are factor type. 

There are 25 different candidate and committee ids. 

Some of the features have to be adjusted before analysis. The contb_receipt_dt feature needs to be a date type. The contbr_zip is an int type but it would make more sense as a factor type.
Some features can be removed because they would not contribute to the analysis. cand_id, file_num, and contbr_st do not add any extra information from an analysis perspective.
It is interesting that while tran_id is a unique id it has less levels than the number of observations.

```{r adjust_features}
# Remove unneeded features
df <- subset(df, select = -c(cand_id, file_num, contbr_st))

# Convert features to correct data type
df$contbr_zip <- factor(df$contbr_zip)
df$contb_receipt_dt <- as.Date(df$contb_receipt_dt, "%d-%b-%y")
```

```{r data_summary}
summary(df)
```

We can see that the contribution receipt dates range 3 years from 2013-12-31 to 2016-12-31. It looks like the committees are matched with the candidate. Also, Hillary Clinton got more than 50% of all the contributions in the dataset. Did she also get more than 50% of the aggregate contribution amount? Most contributions do not name an employer, and a 1/4th of the contributions are from retired people. The majority of receipts had no description.

I think it would useful to know the political party of the candidates.

```{r committee_names}
committees <- levels(df$cand_id)

committees
```



















